Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-12-03 .output/_tex/manuscript.tex
fix fig1 and data variables

\caption{}\label{T_561f9}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.56\% of advanced T-category tumors exhibit midline extension compared
to 6.13\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially accounts
for the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} isolates patients
with lateralized tumors that do not cross the midline, further
demonstrating these relationships. Among early T-category patients with
no ipsilateral nodal involvement (levels I-V), only 1.85\% (2 of 108
patients) show involvement in contralateral level II. This proportion
increases to 6.73\% (14 of 208) if ipsilateral level II is involved, to
14.29\% (10 of 70) if ipsilateral levels II and III are involved, and
further to 39.68\% (25 of 63) for advanced T-category tumors with
ipsilateral levels II and III involved.

@@

For example, we may assess contralateral LNL II involvement prevalence
under early T-category (T0-T2) and no midline extension
(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
observed, with 27 exhibiting contralateral LNL II involvement -- a
prevalence of 7.12\%.

@@

\caption{}\label{T_d0076}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

In this subsection, we illustrate key aspects of the mathematical
framework introduced earlier. The top panel of
figure~\ref{fig-model-midext-evo} shows the prior distribution over
diagnosis times, \(P(t)\). Based on the parameterization, early
T-category tumors are on average diagnosed after 3 time steps, while
advanced T-category tumors are diagnosed later, averaging 4.5 time
steps. This is due to the learned value of \(p_\text{adv.}\), which is
44.98\% for advanced T-category tumors. The tumor's average probability
per time step of growing over the midline, \(p_\epsilon\), was found to
be 8.16\%. Using this value, the conditional probability of midline
extension, \(P(\epsilon \mid t)\), can be computed for a given time step
\(t\) (red line in the top panel of figure~\ref{fig-model-midext-evo}).
The bottom panel visualizes the joint probability \(P(\epsilon, t)\),
showing the likelihood of diagnosis at time \(t\) with specific states
of midline extension and T-category.

@@

Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
ability to account for key risk factors, such as T-category and midline
extension. Consistent with the data, the model predicts that the
prevalence of contralateral LNL II involvement increases from 7.1\% for
early T-category lateralized tumors to 39.17\% for advanced T-category
tumors that cross the midline. Similarly, contralateral LNL III
involvement rises from around 1.55\% for early T-category lateralized
tumors to nearly 14.25\% for advanced T-category midline-extending
tumors.

@@

\begin{itemize}
\tightlist
\item
  A patient with a lateralized early T-category tumor and ipsilateral
  LNL II involvement has a predicted risk of 1.55\% for occult
  contralateral LNL II disease (green histogram).\\
\item
  If the same tumor extends over the midline, the risk increases to
  7.58\% (orange histogram).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  An early T-category tumor crossing the midline with ipsilateral
  involvement of LNLs II and III has a 9.06\% risk of contralateral LNL
  II disease (red histogram).\\
\item
  For the same scenario with an advanced T-category tumor, the risk
  rises to 11.25\% (purple histogram).
\end{itemize}

The degree of ipsilateral involvement also influences risk. For a
midline-extending early T-category tumor, a clinically N0 ipsilateral
neck predicts a risk of 7.58\%, compared to 9.06\% when LNLs II and III
are involved.

@@

\begin{itemize}
\tightlist
\item
  For an advanced T-category tumor extending over the midline and with
  extensive clinical involvement in the ipsilateral LNLs II, III, and
  IV, but a clinically negative contralateral neck, the risk is only
  2.07\% (green histogram).\\
\item
  If contralateral LNL II is clinically involved, the risk for
  contralateral LNL III rises to 4.77\% (blue histogram).\\
\item
  When contralateral LNL II involvement is confirmed by FNA, the risk
  further increases to 6.94\% (red histogram).
\end{itemize}

Even for lateralized tumors, FNA-confirmed involvement in contralateral
LNL II predicts a 5.58\% risk for LNL III involvement (orange
histogram). This highlights the importance of upstream LNL II
involvement in determining the risk for LNL III.

@@

\begin{itemize}
\tightlist
\item
  If contralateral LNL III is also clinically involved, the risk
  increases to 2.81\% (blue histogram).\\
\item
  When contralateral LNL III involvement is confirmed by FNA, the risk
  rises significantly to 5.68\% (orange histogram).
\end{itemize}

@@

For contralateral LNL V, the risk is negligible unless there is severe
contralateral involvement, including LNL IV, confirmed by FNA. In such
extreme cases, the risk increases to 5.49\%.

@@

\caption{}\label{T_7a9be}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
10 \\
early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
61 \\
advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
79 \\
advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\

>>>>

\caption{}\label{T_20ee3}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.6\% of advanced T-category tumors exhibit midline extension compared
to 6.1\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially accounts
for the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} isolates patients
with lateralized tumors that do not cross the midline, further
demonstrating these relationships. Among early T-category patients with
no ipsilateral nodal involvement (levels I-V), only 1.2\% (1 of 86
patients) show involvement in contralateral level II. This proportion
increases to 8.8\% (24 of 272) if ipsilateral level II is involved, to
15.7\% (14 of 89) if ipsilateral levels II and III are involved, and
further to 22.2\% (12 of 54) for advanced T-category tumors with
ipsilateral levels II and III involved.

@@

For example, we may assess contralateral LNL II involvement prevalence
under early T-category (T0-T2) and no midline extension
(\(\epsilon=\texttt{False}\)). In the data, 379 such patients were
observed, with 27 exhibiting contralateral LNL II involvement -- a
prevalence of 7.1\%.

@@

\caption{}\label{T_651fd}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

In this subsection, we illustrate key aspects of the mathematical
framework introduced earlier. The top panel of
figure~\ref{fig-model-midext-evo} shows the prior distribution over
diagnosis times, \(P(t)\). Based on the parameterization, early
T-category tumors are on average diagnosed after 3 time steps, while
advanced T-category tumors are diagnosed later, averaging 4.5 time
steps. This is due to the learned value of \(p_\text{adv.}\), which is
45\% for advanced T-category tumors. The tumor's average probability per
time step of growing over the midline, \(p_\epsilon\), was found to be
8.2\%. Using this value, the conditional probability of midline
extension, \(P(\epsilon \mid t)\), can be computed for a given time step
\(t\) (red line in the top panel of figure~\ref{fig-model-midext-evo}).
The bottom panel visualizes the joint probability \(P(\epsilon, t)\),
showing the likelihood of diagnosis at time \(t\) with specific states
of midline extension and T-category.

@@

Figure~\ref{fig-model-prevalences-overall} demonstrates the model's
ability to account for key risk factors, such as T-category and midline
extension. Consistent with the data, the model predicts that the
prevalence of contralateral LNL II involvement increases from 7.1\% for
early T-category lateralized tumors to 39.2\% for advanced T-category
tumors that cross the midline. Similarly, contralateral LNL III
involvement rises from around 1.5\% for early T-category lateralized
tumors to nearly 14.2\% for advanced T-category midline-extending
tumors.

@@

\begin{itemize}
\tightlist
\item
  A patient with a lateralized early T-category tumor and ipsilateral
  LNL II involvement has a predicted risk of 1.6\% for occult
  contralateral LNL II disease (green histogram).\\
\item
  If the same tumor extends over the midline, the risk increases to
  7.6\% (orange histogram).
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  An early T-category tumor crossing the midline with ipsilateral
  involvement of LNLs II and III has a 9.1\% risk of contralateral LNL
  II disease (red histogram).\\
\item
  For the same scenario with an advanced T-category tumor, the risk
  rises to 11.3\% (purple histogram).
\end{itemize}

The degree of ipsilateral involvement also influences risk. For a
midline-extending early T-category tumor, a clinically N0 ipsilateral
neck predicts a risk of 7.6\%, compared to 9.1\% when LNLs II and III
are involved.

@@

\begin{itemize}
\tightlist
\item
  For an advanced T-category tumor extending over the midline and with
  extensive clinical involvement in the ipsilateral LNLs II, III, and
  IV, but a clinically negative contralateral neck, the risk is only
  2.1\% (green histogram).\\
\item
  If contralateral LNL II is clinically involved, the risk for
  contralateral LNL III rises to 4.8\% (blue histogram).\\
\item
  When contralateral LNL II involvement is confirmed by FNA, the risk
  further increases to 6.9\% (red histogram).
\end{itemize}

Even for lateralized tumors, FNA-confirmed involvement in contralateral
LNL II predicts a 5.6\% risk for LNL III involvement (orange histogram).
This highlights the importance of upstream LNL II involvement in
determining the risk for LNL III.

@@

\begin{itemize}
\tightlist
\item
  If contralateral LNL III is also clinically involved, the risk
  increases to 2.8\% (blue histogram).\\
\item
  When contralateral LNL III involvement is confirmed by FNA, the risk
  rises significantly to 5.7\% (orange histogram).
\end{itemize}

@@

For contralateral LNL V, the risk is negligible unless there is severe
contralateral involvement, including LNL IV, confirmed by FNA. In such
extreme cases, the risk increases to 5.5\%.

@@

\caption{}\label{T_9c098}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
10 \\
early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
61 \\
advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
79 \\
advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\

### 2024-12-03 manuscript.qmd
fix fig1 and data variables

The bottom right panel of @fig-data-strat isolates patients with lateralized tumors that do not cross the midline, further demonstrating these relationships. Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} of {{< var data.early_ipsin0_cII_total >}} patients) show involvement in contralateral level II. This proportion increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} of {{< var data.early_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} of {{< var data.early_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} of {{< var data.late_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.

@@

For example, we may assess contralateral LNL II involvement prevalence under early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$). In the data, {{< var data.early_nomidext_cII_total >}} such patients were observed, with {{< var data.early_nomidext_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_nomidext_cII_percent >}}%.

>>>>

The bottom right panel of @fig-data-strat isolates patients with lateralized tumors that do not cross the midline, further demonstrating these relationships. Among early T-category patients with no ipsilateral nodal involvement (levels I-V), only {{< var data.early_lateral_ipsin0_cII_percent >}}% ({{< var data.early_lateral_ipsin0_cII_match >}} of {{< var data.early_lateral_ipsin0_cII_total >}} patients) show involvement in contralateral level II. This proportion increases to {{< var data.early_lateral_ipsiII_cII_percent >}}% ({{< var data.early_lateral_ipsiII_cII_match >}} of {{< var data.early_lateral_ipsiII_cII_total >}}) if ipsilateral level II is involved, to {{< var data.early_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.early_lateral_ipsiIIandIII_cII_match >}} of {{< var data.early_lateral_ipsiIIandIII_cII_total >}}) if ipsilateral levels II and III are involved, and further to {{< var data.late_lateral_ipsiIIandIII_cII_percent >}}% ({{< var data.late_lateral_ipsiIIandIII_cII_match >}} of {{< var data.late_lateral_ipsiIIandIII_cII_total >}}) for advanced T-category tumors with ipsilateral levels II and III involved.

@@

For example, we may assess contralateral LNL II involvement prevalence under early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$). In the data, {{< var data.early_lateral_cII_total >}} such patients were observed, with {{< var data.early_lateral_cII_match >}} exhibiting contralateral LNL II involvement -- a prevalence of {{< var data.early_lateral_cII_percent >}}%.
